Posterior comparison methods

### 2022-10-31 src/explore_posterior-comparison/posterior-comparison.Rmd
Save samples and import to posterior-comparison
Reading on SBC and PSIS

---
title: "Posterior comparison methods"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background** Multiple approaches for conducting approximate Bayesian inference exist.

@@

@@

@@

@@

@@

>>>>

---
title: "Posterior comparison methods"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
bibliography: citations.bib
abstract: |
  **Background** Multiple approaches for conducting approximate Bayesian inference exist.

@@

# Background

There are lots of ways you can compare posterior distributions.

The simplest is to compare point estimates (such as the mean or mode) or moments (such as the variance).
In some sense, it's the least we should expect from any good approximate Bayesian inference method to be able to recover these summaries.

More challenging is to match the distribution everywhere.
One way to assess this is the Kolmogorov-Smirnov (KS) distance, which gives the maximum distance between two empirical cumulative distribution functions (ECDFs).
It is directly interpretable: a value of 0.05 means that no matter where you compute a tail probability, you're never more than 5% away.

A downside of the KS distance is that it only captures marginal distributions.
This, in some way, favors methods like INLA where we directly approximate marginals.
A further challenge would be to assess agreement between joint distributions, rather than only marginals.

@@

```{r}
samples <- readRDS("depends/samples-m250.rds")
samples_mcmc <- samples$mcmc
samples_gaussian <- samples$gaussian
samples_laplace <- samples$laplace
```

@@

Compute the empirical CDFs, and look for the maximum distance between them.

@@

SBC [@talts2018validating] is a corrected version of @cook2006validation.
Consider the performance of an algorithm over the entire Bayesian joint distribution.
Generate $(\tilde \theta, \tilde y)$ via $\tilde \theta \sim p(\theta)$ then $\tilde y \sim p(y \, | \, \tilde \theta)$.
Let the data averaged posterior (DAP) be
$$
p(\theta) = \int p(y \, | \, \tilde \theta) p(\tilde \theta, \tilde y) \text{d} \tilde \theta \text{d} \tilde y (\#eq:dap)
$$
For any model, the average of any exact posterior expectation with respect to data generated from the Bayesian joint distribution reduces to the corresponding prior expectation.
Any discrepancy between the DAP and the prior indicates some error in the analysis (either implementation or algorithmic).

Consider a sequence of samples from the Bayesian joint distribution
$$
\tilde \theta \sim p(\theta) \\
\tilde y \sim p(y \, | \, \tilde \theta) \\
\{\theta_1, \ldots, \theta_L \} \sim p(\theta \, | \, \tilde y)
$$
Equation \@ref(eq:dap) implies that $\tilde \theta$ and $\{\theta_1, \ldots, \theta_L \}$ should both be distributed according to the prior.
Consequently, for any one-dimensional random variable $f: \Theta \to \mathbb{R}$ the rank statistic of the prior sample relative to the posterior sample
$$
r(\{f(\theta_1), \ldots, f(\theta_L) \}, f(\tilde \theta)) = \sum_{l = 1}^L \mathbb{I}[f(\theta_l) < f(\tilde \theta)] \in [0, L]
$$
will be distributed uniformly across the integers $[0, L]$.
I assume we can use histograms and ECDF difference plots to assess this, and perhaps there will be some additional discussion about how to interpret deviations from uniformity.

# Pareto smoothed importance sampling

Say that we have draws $(\theta_1, \ldots, \theta_S)$ from a proposal distribution $q(\theta)$, then we can estimate $\mathbb{E}_p[h(\theta)]$ by
$$
\frac{\sum_{s = 1}^S h(\theta_s) w_s}{\sum_{s = 1}^S w_s}.
$$
When $w_s = 1$ then this is the direct Monte Carlo estimate, when $w_s = r_s = p(\theta_s, y) / q(\theta)$ then this is importance sampling (IS).
The finite sample performance of IS contains information about how close the proposal distribution is to the target distribution (the true posterior distribution of interest).

Pareto smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes] can improve the estimates from IS by stabilising the importance ratios.
Let
$$
p(y \, | \, \mu, \sigma, k) =
\begin{cases}
\frac{1}{\sigma} \left( 1 + k \left( \frac{y - \mu}{\sigma} \right) \right)^{- \frac{1}{k} - 1}, \quad k \neq 0 \\
\frac{1}{\sigma} \exp \left( \frac{y - \mu}{\sigma} \right), \quad k = 0
\end{cases}
$$
be the generalised Pareto distribution with shape parameter $k$ and location-scale parameter $(\mu, \tau)$.
In PSIS this distribution is fit to the $M$ largest samples of $r_s$ where $M = \min (S/5, 3\sqrt{S})$ (heuristic).
The fitted value $\hat k$ is reported, and these $M$ largest $r_s$ are replaced by their expected value under the Pareto distribution.
All weights are truncated by the raw weight maximum $\max(r_s)$

Can be thought of as a Bayesian version of IS with a prior on the largest importance ratios.
Comment: this is interesting, but seems somewhat suspicious to  me as to why the Bayesian prior would only apply to the largest ratios.
I suppose you could claim that we have genuine prior information that the weights shouldn't be that big.

This procedure is claimed to provide better estimates than plain IS or truncated weights IS.
But it can also be used as a diagnostic.

@@

See [these](http://wittawat.com/assets/talks/ml_meetup_bkk2018.pdf) slides by Wittawat Jitkrittum.
Seth suggests starting with MMD with Gaussian kernel and median heuristic to pick length-scale.
